---
title: "Reproducible Research: Project 2"
author: patrick charles  
output:
    html_document:
        keep_md: true
---

# Storm Data Analysis
- Course: Reproducible Research (repdata-013)
- Project: Course Project 2
- Author: Patrick Charles

# Synopsis

In this report, we'll explore the NOAA Storm Database and answer some
questions about severe weather events. Cumulative health and economic impacts will be evaluated
against the types of events recorded in the dataset.

In the United States, severe weather causes billions of dollars in damage and kills or injures thousands 
of people. In this analysis we'll quantitatively compare the impact of the most damaging types of events.


# Data Processing

## Prerequisite libraries
ggplot2 used for all plots, and tidyr/dplyr pipeline is used for processing.
```{r message=FALSE}
    library(ggplot2)
    library(dplyr)
    library(tidyr)
    library(scales)
```

Historical storm data is obtained from [NOAA/StormData.csv.bz2](https://d396qusza40orc.cloudfront.net/repdata/data/StormData.csv.bz2)

The data is uncompressed, read into a dataframe, damage values scaled/fixed and the results stored as a 
table data frame for further processing and analysis.

```{r readcsv, cache=TRUE}
  # data can be fetched from the source via:
  #   wget https://d396qusza40orc.cloudfront.net/repdata/data/StormData.csv.bz2
  stormfile <- bzfile("StormData.csv.bz2","r")
  df.sd <- read.csv(stormfile) # read the data
  tdf.sd <- tbl_df(df.sd) # convert to a table dataframe for dplyr use
```

```{r processdata}
# adjust property and crop damage values according to magnitude suffix
  tdf.sd$PROPDMG <- ifelse(tdf.sd$PROPDMGEXP == "K", tdf.sd$PROPDMG * 10^3, tdf.sd$PROPDMG)
  tdf.sd$PROPDMG <- ifelse(tdf.sd$PROPDMGEXP == "M", tdf.sd$PROPDMG * 10^6, tdf.sd$PROPDMG)
  tdf.sd$PROPDMG <- ifelse(tdf.sd$PROPDMGEXP == "B", tdf.sd$PROPDMG * 10^9, tdf.sd$PROPDMG)

# str(tdf.sd)
  dim(tdf.sd)
```
The storm data set contains `r dim(tdf.sd)[1]` event observations and measurements
of `r dim(tdf.sd)[2]` different variables.


# Analysis

## Health Impact

To determine which events are most harmful with respect to health,
we'll be looking at the 'FATALITIES' and 'INJURIES' variables

```{r health_analysis}
  tdf.health <- tdf.sd %>% group_by(EVTYPE) %>% # group by event
      summarize(dead=sum(FATALITIES), hurt=sum(INJURIES), 
                total=sum(INJURIES)+sum(FATALITIES))

  # reformat the data and filter events which impacted > threshold people
  health.threshold = 1000
  tdf.health.effects <- gather(tdf.health, variable, value, -EVTYPE, -total)
  tdf.health.worst <- tdf.health.effects %>%
    filter(total > health.threshold) %>% arrange(-total)

  head(tdf.health.worst)
```

The events most harmful to health can be visualized using a stacked
bar chart. Event types are shown which had a cumulative effect on > `r health.threshold` individuals.
```{r health_visual}

  # order the event type factor by magnitude of total effect
  evfactors <- reorder(tdf.health.worst$EVTYPE, tdf.health.worst$total)

  ggplot(tdf.health.worst, aes(x=evfactors, y=value, fill=variable)) +
    geom_bar(stat="identity") + coord_flip() +
    ggtitle("Most Harmful Event Types to Population Health") +
    ylab("Total Number Impacted") + xlab("Event Type")
```

## Economic Impact

To determine which events are most harmful with respect to economic cost,
we'll be looking at the crop and property damage variables.

```{r economic_analysis}
  tdf.econ <- tdf.sd %>% group_by(EVTYPE) %>% # group by event
      summarize(crop=sum(CROPDMG), prop=sum(PROPDMG), 
                total=sum(CROPDMG)+sum(PROPDMG))

  # reformat data and filter events with total economic impact > econ.threshold
  econ.threshold = 3000000000
  tdf.econ.effects <- gather(tdf.econ, variable, value, -EVTYPE, -total)
  tdf.econ.worst <- tdf.econ.effects %>%
    filter(total > econ.threshold) %>% arrange(-total)

  head(tdf.econ.worst)
```

The most economically harmful events can be visualized using a stacked
bar chart. Event types are shown which had a cumulative economic impact > $`r econ.threshold`.
```{r economic_visual}
  # order the event type factor by magnitude of total effect
  evfactors <- reorder(tdf.econ.worst$EVTYPE, tdf.econ.worst$total)

  ggplot(tdf.econ.worst, aes(x=evfactors, y=value, fill=variable)) +
    geom_bar(stat="identity") + coord_flip() + 
    ggtitle("Most Economically Harmful Event Types") +
    ylab("Total $ Impact (Crops and Property)") + xlab("Event Type") + 
    theme(legend.position="none") # suppress legend
```


# Results

- Across the United States, the following types of events are most
harmful to population health: Tornadoes, Heat, Wind, Flooding and Lightning

- Across the United States, the following types of events have
the greatest economic consequences: Floods, Hurricanes/Typhoons, Tornadoes, Storm Surges, and Flash Floods
