---
title: "Reproducible Research: Project 2"
author: patrick charles  
output:
    html_document:
        keep_md: true
---

# Storm Data Analysis
- Course: Reproducible Research (repdata-013)
- Project: Course Project 2
- Author: Patrick Charles

# Synopsis

In this report, we'll explore the NOAA Storm Database and answer some
questions about severe weather events. Cumulative health and economic impacts will be evaluated
against the types of events recorded in the dataset.

The data set analyzed contains all storm events in the United States as entered by NOAA's National Weather Service (NWS) from 1950 through 2011.

Severe weather causes billions of dollars in damage and kills or injures thousands 
of people. In this analysis we'll quantitatively compare the impact of the most damaging types of events.


# Data Processing

## Prerequisite libraries
ggplot2 used for all plots, and tidyr/dplyr pipeline is used for processing.
```{r message=FALSE}
    library(ggplot2)
    library(dplyr)
    library(tidyr)
    library(scales)
```

Historical storm data is obtained from [NOAA/StormData.csv.bz2](https://d396qusza40orc.cloudfront.net/repdata/data/StormData.csv.bz2)

The data is uncompressed, read into a dataframe, damage values scaled and the results stored as a 
table data frame for further processing and analysis.

```{r readcsv, cache=TRUE}
  # data can be fetched from the source via:
  #   wget https://d396qusza40orc.cloudfront.net/repdata/data/StormData.csv.bz2
  stormfile <- bzfile("StormData.csv.bz2","r")
  df.sd <- read.csv(stormfile) # read the data
  tdf.sd <- tbl_df(df.sd) # convert to a table dataframe for dplyr use
```

```{r processdata}
  # conversion function to scale values based on separate magnitude 
  scaleValue <- function(val, exp) {
    switch(exp, "K" = val * 10^3, "M" = val * 10^6, "G" = val * 10^9, val)
  }
  scaleVector = Vectorize(scaleValue) # vectorized version of the scale function

  # scale damage values by the multiplier stored in the separate 'exp' variable
  tdf.sd$PROPDMG <- scaleVector(tdf.sd$PROPDMG, as.character(tdf.sd$PROPDMGEXP))
  tdf.sd$CROPDMG <- scaleVector(tdf.sd$CROPDMG, as.character(tdf.sd$CROPDMGEXP))

  # convert strings to dates
  tdf.sd$BGN_DATE <- as.Date(tdf.sd$BGN_DATE, "%m/%d/%Y") # convert event start dates
  tdf.sd$END_DATE <- as.Date(tdf.sd$END_DATE, "%m/%d/%Y") # convert event end dates

  first.date <- min(tdf.sd$BGN_DATE) # find the date of the earliest event
  last.date <- max(tdf.sd$BGN_DATE) # find the date of the latest event
  dim(tdf.sd) # query the total number of observations
```
The storm data set contains `r dim(tdf.sd)[1]` event observations and measurements
of `r dim(tdf.sd)[2]` different variables.

The storm data set covers events occurring during the period from `r first.date` to `r last.date`.


# Analysis

## Health Impact

To determine which events are most harmful with respect to health,
we'll be looking at the 'FATALITIES' and 'INJURIES' variables

```{r health_analysis}
  tdf.health <- tdf.sd %>% group_by(EVTYPE) %>% # group by event
      summarize(Killed=sum(FATALITIES), Injured=sum(INJURIES), 
                total=sum(INJURIES)+sum(FATALITIES))

  # reformat the data and filter events which impacted > threshold people
  health.threshold = 1000
  tdf.health.effects <- gather(tdf.health, impact, value, -EVTYPE, -total)
  tdf.health.worst <- tdf.health.effects %>%
    filter(total > health.threshold) %>% arrange(-total)

  head(tdf.health.worst)
```

The events most harmful to health can be visualized using a stacked
bar chart. Event types are shown which had a cumulative effect on > `r health.threshold` individuals.
```{r health_visual}

  # order the event type factor by magnitude of total effect
  evfactors <- reorder(tdf.health.worst$EVTYPE, tdf.health.worst$total)

  ggplot(tdf.health.worst, aes(x=evfactors, y=value, fill=impact)) +
    geom_bar(stat="identity") + coord_flip() +
    ggtitle("Most Harmful Event Types to Population Health") +
    ylab("Total Number Impacted (from 1950 through 2011)") + xlab("Event Type")
```

## Economic Impact

To determine which events are most harmful with respect to economic cost,
we'll be looking at the crop and property damage variables.

```{r economic_analysis}
  tdf.econ <- tdf.sd %>% group_by(EVTYPE) %>% # group by event
      summarize(Crop=sum(CROPDMG), Property=sum(PROPDMG), 
                total=sum(CROPDMG)+sum(PROPDMG))

  # reformat data and filter events with total economic impact > econ.threshold
  econ.threshold = 3000000000
  tdf.econ.effects <- gather(tdf.econ, impact, value, -EVTYPE, -total)
  tdf.econ.worst <- tdf.econ.effects %>%
    filter(total > econ.threshold) %>% arrange(-total)

  head(tdf.econ.worst)
```

The most economically harmful events can be visualized using a stacked
bar chart. Event types are shown which had a cumulative economic impact > $`r econ.threshold`.
```{r economic_visual}
  # order the event type factor by magnitude of total effect
  evfactors <- reorder(tdf.econ.worst$EVTYPE, tdf.econ.worst$total)

  ggplot(tdf.econ.worst, aes(x=evfactors, y=value, fill=impact)) +
    geom_bar(stat="identity") + coord_flip() + 
    ggtitle("Most Economically Harmful Event Types") +
    ylab("Total $ Impact (from 1950 through 2011)") + xlab("Event Type")
```


# Results

- Across the United States, the following types of events are most
harmful to population health: Tornado, Excessive Heat, Wind, Flood and Lightning

- Across the United States, the following types of events have
the greatest economic consequences: Tornado, Flood, Hail, Flash Flood, and Drought

- These results assume that the EVTYPE variable contains distinct and unique event descriptors. In the provided data, this is not the case for all events. Events sometimes use inconsistent labels (e.g. "COLD", "COLD WAVE", "RECORD COLD") to describe the same or similar events. Also, many of the observations confound multiple event types (e.g. "SNOW AND COLD"", "FOG AND COLD"). A more precise analysis of the events would systematically address these inconsistencies.


```{r deathratio, echo=FALSE}
# some extra stats, of interest
  deathratio.threshold = 0.2
  individualMin = 100
  tdf.deathratio <- tdf.health %>% 
    mutate(DeathRatio=ifelse(total != 0, Killed/total, NA))

  tdf.deathratio.worst <- tdf.deathratio %>%
    filter(DeathRatio > deathratio.threshold) %>% filter(total > individualMin) %>% 
    arrange(-DeathRatio)

  evfactors <- reorder(tdf.deathratio.worst$EVTYPE, tdf.deathratio.worst$DeathRatio)

  #ggplot(tdf.deathratio.worst, aes(x=evfactors, y=DeathRatio)) +
  #  geom_bar(stat="identity") + coord_flip() +
  #  ggtitle("Deadliest") +
  #  ylab("Ratio of Killed to Injured (from 1950 through 2011)") + xlab("Event Type")
```
